\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-30 syntax.tex
Fix the introduction of sealing Fix meta-theory section.

Before we dive in the details of our metatheory, we have to mention that, the challenge of incorporating family polymorphism with inductive type is quite different in metatheory than in surface syntax because of de Bruijn indices. (Using surface synax), given a linkage $\{a~:~\bW \sigma; b_1 : a \to T;..\}$ with a bunch of $b_i$ \textit{using} $a$. The tension we need to solve here is between exhaustiveness checking and inheritance---on one hand, if we want to make sure $b_i$ is pattern matching correctly, then we have to let $b_i$ term construct in the context of $\bW\sigma$; on the hand, if we want to let $b_i$ to be inherited into children family with extended inductive type $a~:~\bW \sigma'$, then the old inductive type $a~:~\bW \sigma$ in the context of $b_i$ is stopping that. Our metatheory formulation of linkage is mainly trying to solve this tension.

$\goodSig{\Gamma}{\sigma}{n}$ means $\sigma$ is a well-typed signature
for linkages in the context $\Gamma$ with $n$ fields. We omit the
naming/label information for fields in the meta\-theory. We can use $\cL$
to retrieve the corresponding types of linkage of a given signature, and
$\cC$ to get the compiled linkage type (an existential type).
$\cL\sigma$ corresponds to the type of a family term, but $\cC\sigma$, accoding to \ruleref{Compile}, is a deeply nested dependent pair. Then $\cCt \sigma$ provide the concrete module of type $\cC\sigma$ given a linkage of type $\cL\sigma$.

The reason we need this additional $\cC\sigma$ here is because, the rule of linkages hinders its ability to do abstraction. Concretely speaking, (using surface syntax) we don't have \\ $\goodTerm{\Gamma, \cL(\{a~:~\bW \sigma; b~:~a \to T\})}{h}{\cL(\{a~:~\cU; b~:~a \to T\})}$\footnote{If we had one, consider the case $T$ is bottom type and $\sigma$ is empty record and thus $a$ is initially bottom type. Now we further override $h.a$ using unit type we can derive inconsistency}. This term is important for us to abstract the concrete inductive type so that the later field can depend on the abstract interface. On the other hand, we do have $\goodTerm{\Gamma, (\Sigma a : \bW\sigma, a \to T) }{h}{(\Sigma a : \cU, a \to T)}$ because of the rule for existential (also existential type just don't allow overriding). Thus to allow abstraction happens, we need  $\cC\sigma$ to compile a linkage to module and we also need \ruleref{Lnkg Compile} to specify the compilation from linkage to module.

@@

The signature (for the linkage) and the linkage rules are quite similar to
the signature for inductive types.  They are both constructed from the
empty stuff, and respect substitution lemmas. We have
\ruleref{Sig Proj}, \ruleref{Lnkg Proj} projection for each component
due to harmony~\cite{pfenning2009lecture}, so we have local soundness
and local completeness ($\beta,\eta$ rules). We omit them and the
substitution laws for either field addition and projection. For
$\mu^+,\nu^+$ rules, we will omit $\{f\}$ if possible.\footnote{This
omission can be supported by Agda using Implicit Arguments}

Since our recursor handlers are closely related to linkages, we can finally talk about the rules of recursor now.

Now, one missing piece for linkage is the sealing judgement, and both of the $\mu^+, \nu^+$ rules relies on that---then what is this sealing judgement for?

@@

The motivation for sealing judgement is still about the important term we mentioned earlier: it is acting like syntactic sugar
helpful for inheritance---due to the presence of inductive type in the
context, not everything is inheritable. For example, $ X_1 : \bW
\sigma_1 \vdash t : T$ in the parent family will generally not be able
to be inherited to the children once we override the inductive type $\bW
\sigma_1$ with $\bW \sigma_2$. In
other words, there is no way to ``transplant'' $t$ into the context $X_2
: \bW \sigma_2$.
This makes all the fields after this inductive type
non-inheritable---which shouldn't be the case because a lot of fields
might not be directly relying on this concrete $\bW \sigma_1$.

Sealing\YZ{It occurs to me that you use the word "sealing" to mean different (though tangentially related) things. Previously you used it as in "sealed families".}\EDJreply{They are actually the same thing. This sealing in metatheory is much stronger than that of the implemented plugin. Downside of the strong sealing is that if the user really use this version of sealing, the user will have to annotate what the sealed family type is. Now in the plugin, the sealed family type is almost fixed (directly inferred using a fixed rules). So basically this complete version of sealing (in metatheory) can deduce that of seal (in plugin implementation) } is used to exploit this fact---with sealing, given $X_1 : \bW
\sigma_1$, we can add well-typed terms $X_1 : \cU \vdash t : T$, into
the parent family. Later when we want to inherit $t$,
because $t$ only consider $X_1$ as an arbitrary type, inheritance is
possible. This sealing is the general version of the sealed family implemented in the plugin. Our plugin will deduce a sealed ``interface'' for a sealed family using a fixed set of rules but here sealing in metatheory can be arbitrary local morphism.

Of course, any fields using "Wsup" and "Wrec" directly relies on the
$\bW(\sigma)$ (can be spotted from \ruleref{Ind-Term} and
\ruleref{Rec-Constr}) and thus non-inheritable---in these cases, "Wsup"
can be understood as concrete constructors and "Wrec" are
\textit{exhaustiveness checking}. These fields together with inductive
types are the ones sealing is responsible for hiding. After compilation
of linkage types into existential types, we can easily hide the
\textit{concrete} definition of the inductive type while exposing all
the fields using "Wsup" and "Wrec" like an abstract interface.
Apparently, future fields relying on this abstract interface will have no
problem on inheritance.
Of course, since "Wsup" and "Wrec" directly rely on the concrete inductive definition, we will have to override these fields whenever inductive type is overridden. For "Wsup" we only need to override each constructor and for "Wrec" it is just doing exhaustiveness checking inside each children family. The recursor handlers however, can be safely inherited.
\YZ{I suppose the terminology 'sealing' is a reference to ML modules? In
ML, sealing has the connotation of hiding things behind an existential
type. So using the same word may cause the reader to have false
presumptions about what is going on here.}\EDJreply{You are right. Our "Sealing" is not hiding a particular field into an existential type, but our seal is make several fields together into an "existential type". Basically it hides the concrete definition of inductive type so that future fields doesn't have to be defined upon that.}\EDJreply{I will resolve this comment once I add the throughout example and you find it clear---because I haven't changed any text}

There are also cases we don't need to seal anything (i.e., doing
exhaustiveness checking when using recursor), and thus the "Seal-Id" is
used. For notation clarity, we will sometimes use $"sf'"$ to emphasize
that we consider that seal $s$ as a term $("sf'" \ s)$.

With the definition of sealing judgement in mind, we can see
\ruleref{Sig Add} and \ruleref{Lnkg Add} work exactly as the
implementation. In the implementation, we will have "self__" all around
when defining new fields in our plugin, and this "self__" is exactly the
$A$ in the context of the two fields---especially when we don't do any
sealing, this $A = \cC \sigma$, which is exactly how the implementation
exposes "self__" as a module parameter.

@@

Finally we introduce the inheritance judgement. Inheritance are
judgement so naturally second class citizen data (i.e. not something
function can return and cannot be assigned to variables), and it looks
very much like a function that transform a given linkage.

@@

Override "inhov" is even more special because in mundane OO programming language, we usually require that the overriding term has the same type/interface as the overriden one.  Here we don't since the very reason we want the same type/interface is that we hope other inherited fields can use this overridden/late-binding field without breaking abstraction, and this reason is already managed by the assumption of "inhinh", the upcasting proof $\uparrow^s$. Thus "inhov" is quite like "inhext", where overriding is just throwing away one parent field. This difference also show one of the distinct feature of our vanilla family inheritance compared to the mainstream OO inheritance---family inheritance don't need to introduce the concept of ``subclassing'' between families and thus no bounded polymorphism yet\footnote{Of course once trait and interface comes in, we can have bounded polymorphism back}, and code reuse is mainly achieved by late binding. Without ``subclassing'', there is no way of confusing inheritance and subtyping.

>>>>

$\goodSig{\Gamma}{\sigma}{n}$ means $\sigma$ is a well-typed signature
for linkages in the context $\Gamma$ with $n$ fields. We omit the
naming/label information for fields in the meta\-theory. The signature (for the linkage) and the linkage rules are quite similar to
the signature for inductive types.  They are both constructed from the
empty stuff, and respect substitution lemmas. We have
\ruleref{Sig Proj}, \ruleref{Lnkg Proj} projection for each component
due to harmony~\cite{pfenning2009lecture}, so we have local soundness
and local completeness ($\beta,\eta$ rules). We omit them and the
substitution laws for either field addition and projection. For
$\mu^+,\nu^+$ rules, we will omit the sealing $\{f\}$ if possible.\footnote{This
omission can be supported by Agda using Implicit Arguments}

A side note: before we dive in more details of our metatheory for linkage (especially the meaning of sealing and $\cC$), we have to mention that, the challenge of incorporating family polymorphism with inductive type mentioned in \ref{chg:extensible-inductive-type} mutates in metatheory because of de Bruijn indices. (Using surface synax), consider the example of a linkage $\{a~:~\bW \sigma; b_1 : a \to T;..\}$ with a bunch of $b_i$ \textit{using} $a$. The tension we need to solve here is between \textit{exhaustiveness checking} and \textit{inheritance (code reuse)}---on one hand, if we want to make sure $b_i$ is pattern matching correctly, then we have to construct $b_i$ term in the context of $\bW\sigma$; on the hand, if we want to let $b_i$ to be inherited into children family with extended inductive type $a~:~\bW \sigma'$, then the old inductive type $a~:~\bW \sigma$ in the context of $b_i$ is stopping that. Our metatheory formulation of linkage is mainly trying to solve this tension.

We use $\cL$ to retrieve the corresponding types of linkage of a given signature, and $\cC$ to get the compiled linkage type (an existential type).
$\cL\sigma$ corresponds to the type of a family term, but $\cC\sigma$, according to \ruleref{Compile}, is a deeply nested dependent pair. Then $\cCt \sigma$ provide the concrete module of type $\cC\sigma$ given a linkage of type $\cL\sigma$.

The reason we need this additional type construct of $\cC\sigma$ here is to support \textit{code reuse}, as the rule of linkages hinders its ability to do abstraction. Concretely speaking, (using surface syntax) we don't have \\ $\goodTerm{\Gamma, \cL(\{a~:~\bW \sigma; b~:~a \to T\})}{h}{\cL(\{a~:~\cU; b~:~a \to T\})}$\footnote{If we had one, consider the case $T$ is bottom type and $\sigma$ is empty record and thus $a$ is initially bottom type. Now we further override $h.a$ using unit type we can derive inconsistency}. This term is important for us to abstract the concrete inductive type so that the later field can depend on the abstract interface and achieve \textit{code reuse} reuse. On the other hand, we do have $\goodTerm{\Gamma, (\Sigma a : \bW\sigma, a \to T) }{h}{(\Sigma a : \cU, a \to T)}$ because of the rule for existential (also existential type just don't allow overriding). Thus to allow abstraction happens, we need  $\cC\sigma$ to compile a linkage to module as a part of the context and we also need \ruleref{Lnkg Compile} to specify the compilation from linkage to module.

@@

However, $\cC$ can only make sure we have $\cL(\{a~:~\bW \sigma; b~:~a \to T\}) \to (\Sigma a:\bW \sigma, a \to T)$, we still need $(\Sigma a:\bW \sigma, a \to T) \to (\Sigma a:\cU \sigma, a \to T)$ to complete this shifting of the context. This is sealing's job.

@@

Sealing judgement is still about the important term $h$ we mentioned earlier. Thanks to $\cC$, a newly added term $t_{new}$ is now in the context of $\cC\{a : \bW \sigma; ..\}$. But if we want $t_{new}$ to be inherited(reused) when $\sigma$ is extended, we would like $t_{new}$ to be in the context  of $\cC\{a : \cU; ..\}$ instead. The sealing is responsible for this because we have a sealing $\goodTerm{\Gamma, \cC\{a : \bW \sigma; ..\}}{s}{\cC\{a : \cU; ..\}}$ = $\goodTerm{\Gamma, (\Sigma a : \bW \sigma,..)}{s}{(\Sigma a : \cU,..)}$. Then we can construct $t_{new}$ in the context of  $(\Sigma a : \cU,..)$, ready for inheritance. Apparently $t_{new}$ in this case cannot be doing \textit{exhaustiveness checking}, but it can freely use any \textit{exhaustiveness-checked} recursor. \YZ{It occurs to me that you use the word "sealing" to mean different (though tangentially related) things. Previously you used it as in "sealed families".}\EDJreply{They are actually the same thing. This sealing in metatheory is much stronger than that of the implemented plugin. Downside of the strong sealing is that if the user really use this version of sealing, the user will have to annotate what the sealed family type is. Now in the plugin, the sealed family type is almost fixed (directly inferred using a fixed rules). So basically this complete version of sealing (in metatheory) can deduce that of seal (in plugin implementation) }

% For example, $ X_1 : \bW
% \sigma_1 \vdash t : T$ in the parent family will generally not be able
% to be inherited to the children once we override the inductive type $\bW
% \sigma_1$ with $\bW \sigma_2$. 
% In
% other words, there is no way to ``transplant'' $t$ into the context $X_2
% : \bW \sigma_2$.
% This makes all the fields after this inductive type
% non-inheritable---which shouldn't be the case because a lot of fields
% might not be directly relying on this concrete $\bW \sigma_1$.

% Sealing is used to exploit this fact---with sealing, given $X_1 : \bW
% \sigma_1$, we can add well-typed terms $X_1 : \cU \vdash t : T$, into
% the parent family. Later when we want to inherit $t$,
% because $t$ only consider $X_1$ as an arbitrary type, inheritance is
% possible.

This sealing is the general version of the sealed family implemented in the plugin. Our plugin will deduce a sealed ``interface'' for a sealed family using a fixed set of rules but here sealing in metatheory can be arbitrary local morphism. With the definition of sealing judgement in mind, we can see
\ruleref{Sig Add} and \ruleref{Lnkg Add} work exactly as the
implementation. In the implementation, we will have "self__" all around
when defining new fields in our plugin, and this "self__" is exactly the
$A$ in the context of the two fields---especially when we don't do any
sealing, this $A = \cC \sigma$, which is exactly how the implementation
exposes "self__" as a module parameter.

In summary, with the help of both sealing and $\cC$, we can shift the context from $\cL(\{a~:~\bW \sigma; ..\})$ to $\Sigma a : \cU,..$ and thus achieve code reuse. However, this is just allow possible \textit{code reuse}, we still need to solve the mutated tension in \ref{chg:extensible-inductive-type}. To make things concrete, we need to introduce the rules for recursors.

% Since our recursor handlers are closely related to linkages, we can finally talk about the rules of recursor now.

Again, we still use \textit{decoupling} of the syntax of the case handlers and exhaustiveness checking to resolve \ref{chg:extensible-inductive-type}. The linkage is the part of defining case handlers and thus inheritable like other family.
While any fields using "Wsup" and "Wrec" directly relies on the
$\bW(\sigma)$ (can be spotted from \ruleref{Ind-Term} and
\ruleref{Rec-Constr}) and thus non-inheritable---in these cases, "Wsup"
can be understood as concrete constructors and "Wrec" are
\textit{exhaustiveness checking}. These fields together with inductive
types are the ones sealing is responsible for hiding. After compilation
of linkage types into existential types, we can easily hide the
\textit{concrete} definition of the inductive type while exposing all
the fields using "Wsup" and "Wrec" like an abstract interface.
Apparently, future fields relying on this abstract interface will have no
problem on inheritance.
Of course, since "Wsup" and "Wrec" directly rely on the concrete inductive definition, we will have to override these fields whenever inductive type is overridden. For "Wsup" we only need to override each constructor and for "Wrec" it is just doing exhaustiveness checking inside each children family. The recursor handlers however, can be safely inherited.
\YZ{I suppose the terminology 'sealing' is a reference to ML modules? In
ML, sealing has the connotation of hiding things behind an existential
type. So using the same word may cause the reader to have false
presumptions about what is going on here.}\EDJreply{You are right. Our "Sealing" is not hiding a particular field into an existential type, but our seal is make several fields together into an "existential type". Basically it hides the concrete definition of inductive type so that future fields doesn't have to be defined upon that.}\EDJreply{I will resolve this comment once I add the throughout example and you find it clear---because I haven't changed any text}

There are also cases we don't need to seal anything (i.e., for doing
exhaustiveness checking when using recursor), and thus the "Seal-Id" is
used. For notation clarity, we will sometimes use $"sf'"$ to emphasize
that we consider that seal $s$ as a term $("sf'" \ s)$.

@@

Inheritance are
judgement so naturally second class citizen data (i.e. not something
function can return and cannot be assigned to variables), and it looks
very much like a function that transform a given linkage.

@@

Override "inhov" is even more special because in mundane OO programming language, we usually require that the overriding term has the same type/interface as the overriden one.  Here we don't since the very reason we want the same type/interface is that we hope other inherited fields can use this overridden/late-binding field without breaking abstraction, and this reason is already managed by the assumption of "inhinh", the upcasting proof $\uparrow^s$. Thus "inhov" is quite like "inhext", where overriding is just throwing away one parent field. This difference also show one of the distinct feature of our vanilla family inheritance compared to the mainstream OO inheritance---family inheritance don't need to introduce the concept of ``subclassing'' between families and thus no bounded polymorphism yet\footnote{Of course once trait and interface comes in, we can have bounded polymorphism back}, and code reuse is mainly achieved by late binding. Without ``subclassing'', there is no way of confusing inheritance and subtyping. Note that, "inhov" is also responsible for ``extending'' inductive type---the extended inductive type directly override the old ones when using metatheory to program, just like how plugin implementation works.
